---
title: "Data Report 2"
author: "Final Assignment"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(pcaMethods)
```


## Part 1: PCA
```{r, echo = FALSE}
proteins <- read.csv("C:/Users/Ben/Downloads/protein_data.csv")
t_proteins <- t(proteins[,-1])
my_pca <- pca(t_proteins, method = "ppca")
```

I chose to use the "pca" method from pcaMethods because this implementation can handle missing values, while prcomp cannot. There are many missing values in this data set, and I would rather use a method that can natively deal with them over manually removing and losing data. 

```{r, echo = FALSE, fig.cap="Figure 1: First two principal components from PCA explain a total of 81.3% of total variance"}
pca_df <- merge(t_proteins, scores(my_pca), by = 0)
pca_df$Type = ifelse(grepl("Core", pca_df$Row.names), "Core", "STB")
pca_df <- pca_df %>% relocate(Type)
pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Type)) + 
  geom_point() +
  stat_ellipse() +
  xlab("PC1 (64.4%)") + 
  ylab("PC2 (16.9%)")

pca_plot
```

```{r, echo = FALSE}
loads <- data.frame(loadings(my_pca))
largest <- order(abs(loads$PC1), decreasing = T)

head(largest)
```
Protein 806 has the greatest absolute value loading on PC1

## Part 2: Survival Analysis

We begin with the BrainCancer data set, which is part of the ISLR2 package.

```{r, echo = FALSE, warning = FALSE}
library(ISLR2)
data(BrainCancer)

```

The rows index the 88 patients, while the columns contain the 8 predictors.

```{r}
names(BrainCancer)
```

1. How many female and how many male patients make up this dataset?
```{r, echo = FALSE}
table(BrainCancer$sex)
```

There are 45 female and 43 male patients in this dataset

2. Using diagnosis, how many diagnosis types are there?
```{r, echo = FALSE}
table(BrainCancer$diagnosis)
```

There are three distinct diagnosis types, plus an "other" category


```{r}
table(BrainCancer$status)
```

Before beginning an analysis, it is important to know how the status variable
has been coded. Most software, including R, uses the convention that
status = 1 indicates an uncensored observation, and status = 0 indicates
a censored observation. But some scientists might use the opposite coding.
For the BrainCancer data set 35 patients died before the end of the study.


Based on the status table, it appears that this data set uses 1 to represent a censored observation


To begin the analysis, we create a Kaplan-Meier survival curve, using the survfit() function within the R survival library. Here time corresponds to Y and status corresponds to $\delta$.

```{r, echo = FALSE, warning = FALSE, fig.cap = "Figure 2: Kaplan-Meier curve, depicting probability of survival over time for all patients"}
# make sure to install the survival package #
library(survival)
fit.surv <- survfit(Surv(time, status) ~ 1, data = BrainCancer)

# the survival package uses base R to plot, so you may see some unusual commands.
# there is a package called ggsurvplot that you can explore for ggplot
plot(fit.surv , xlab = "Months",
ylab = "Estimated Probability of Survival")
```

Next we create Kaplan-Meier survival curves that are stratified by sex

```{r, echo = FALSE, warning = FALSE, fig.cap = "Figure 3: Kaplan-Meier curve, depicting probability of survival over time for all patients stratified by sex"}
fit.sex <- survfit(Surv(time, status) ~ sex, data = BrainCancer)
plot(fit.sex , xlab = "Months",
ylab = "Estimated Probability of Survival", col = c(2,4))
legend("bottomleft", levels(BrainCancer$sex), col = c(2,4), lty = 1)
```

We can perform a log-rank test to compare
the survival of males to females, using the survdiff() function. 

2. Based on the computed p-value and a null hypothesis of no difference in survival, what do you conclude about this data?

```{r, echo = FALSE, warning = FALSE}
logrank.test <- survdiff(Surv(time, status) ~ sex, data = BrainCancer)
logrank.test$pvalue
```


With a p-value of 0.23, we fail to reject the null hypothesis. There is insufficient evidence to conclude there is any difference is survival between sexes


Next, we fit Cox proportional hazards models using the coxph() function. To begin, we consider a model that uses sex as the only predictor. coxph()

```{r, echo = FALSE, warning = FALSE}
fit.cox <- coxph(Surv(time, status) ~ sex, data = BrainCancer)
summary(fit.cox)
```

3. Is there a difference in males and females based on the Cox regression (Likelihood ratio test)? Explain your answer.


The Cox regression test yields similar results to the log rank test -- with sexMale having a p-value of .23, we again fail to reject null.


Now we fit a model that makes use of additional predictors.

```{r, echo = FALSE, warning = FALSE}
fit.all <- coxph(
Surv(time, status) ~ sex + diagnosis + loc + ki + gtv +
stereo, data = BrainCancer)
summary(fit.all)
```

4. Which other predictors are significant for survival?


There are two predictors that achieve significance: a diagnosis of HG (High Grade) Glioma, and "ki". KI represents the Karnovsky index, an integer ranging from 0-100 that represents a patient's degree of functional impairment. A score of 100 represents no evidence of disease, and 0 represents death.


Finally, we plot survival curves for each diagnosis category, adjusting for
the other predictors. To make these plots, we set the values of the other
predictors equal to the mean for quantitative variables, and the modal value
for factors. We first create a data frame with four rows, one for each level
of diagnosis. The survfit() function will produce a curve for each of the
rows in this data frame, and one call to plot() will display them all in the
same plot.

```{r, echo = FALSE, warning = FALSE, fig.cap = "Figure 4: Probability of survival stratified by diagnosis category"}
modaldata <- data.frame(
diagnosis = levels(BrainCancer$diagnosis),
sex = rep("Female", 4),
loc = rep("Supratentorial", 4),
ki = rep(mean(BrainCancer$ki), 4),
gtv = rep(mean(BrainCancer$gtv), 4),
stereo = rep("SRT", 4)
)

survplots <- survfit(fit.all, newdata = modaldata)
plot(survplots , xlab = "Months",
ylab = "Survival Probability", col = 2:5)
legend("bottomleft", levels(BrainCancer$diagnosis), col = 2:5, lty = 1)
```
